---
title: "Untitled"
format: pdf
pdf-engine: lualatex
editor: visual
mainfont: Calibri
monofont: PT Mono
always_allow_html: yes
header-includes:
   \usepackage[dvipsnames]{xcolor}
   \definecolor{darkblue}{rgb}{0.0, 0.0, 0.55}
   \definecolor{ivory}{rgb}{1.0, 1.0, 0.94}
---

```{r, include=FALSE}
# This will allow to use different font sizes inside code chunks
# Won't be included in the report
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

### Preprocessing reads

##### Preparing reads

Dorado doesn't support demultiplexing of dual indexes on 5' and 3' ends. Additionally, ligated library reads can be either orientation. Most straightforward approach is to solve this, is to utilise flanking adapter sequences and cutadapt.

Extracting forward reads to fastq file can be performed with following command

*cutadapt -g t2f...rc(t2r) -O 13 --trimmed-only -m 1300 -M 1650 -o forward_out.fastq.gz raw_reads.fastq.gz*

Extracting reverse reads

*cutadapt -g t2r...rc(t2f) -O 13 -m 1300 -M 1650 --trimmed-only -o reverse_out.fastq.gz raw_reads.fastq.gz*

**Note!** e=0, minimum (m) and maximum size limits reduce chance of misaligned reads

Next step is to reverse complement reverse read file and join all reads to single file. This will make life easier later in the data analysis workflow.

*seqkit seq -rp --seq-type DNA reverse_out.fastq.gz \| gzip \>reverse_comp.fastq.gz*

Next step is to merge two files together

*cat forward_out.fastq.gz reverse_comp.fastq.gz \>nanopore_reads.fastq.gz*

##### Demultiplexing reads

Prepare list of barcodes as a fasta file

Use cutadapt to demux fastq file. In example output files are written to demuxed subdirectory

*cutadapt -e 0 -O 12 -g file:barcodes.fasta -o "demuxed/{name}.fastq.gz" input.fastq.gz*

##### Trimming PCR amplification primers

You can use cutadapt and bash scripts to trim forward and revers PCR primers from demultiplexed sequence files.

### Importing set1

Loading libraries

```{r, warning=FALSE, message=FALSE}
library(dada2);packageVersion("dada2")
library(knitr);packageVersion("knitr")
library(Biostrings);packageVersion("Biostrings")
library(DECIPHER);packageVersion("DECIPHER")
library(phyloseq);packageVersion("phyloseq")
library(tidyverse);packageVersion("tidyverse")
#library(ips);packageVersion("ips")
#library(ape);packageVersion("ape")
library(kableExtra);packageVersion("kableExtra")
#library(patchwork);packageVersion("patchwork")
#library(mia);parckageVersion("mia")
```

```{r, warning=FALSE,message=FALSE, size="small"}
# Path variables
path <- "data/processed/set1"
training <- "~/feature_classifiers/SILVA_SSU_r138_2019.RData"
#meta_file <- "~/pr_eng/hanna/metadata.tsv"
#exportloc <- "~/pr_eng/hanna/hanna_results/"
# Variables: truncation length, phix (Illumina)
truncation <- 1400
# Set name of first column in your metadata file
#meta_1stcol <- "sampleid"
#Creates results directory
#dir.create(exportloc)
```

```{r, warning=FALSE, message=FALSE, size="small"}
#List files inside directory
list.files(path)
# Forward fastq filenames have format: SAMPLENAME_R1_001.fastq
fnFs <- sort(list.files(path, pattern="_trimmed_all.fastq.gz", full.names = TRUE))
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

\newpage

Plotting sequence quality profile for sample pairs

```{r, warning=FALSE, message=FALSE, size="small"}
# Base quality plot
prI <- plotQualityProfile(fnFs[1:2])
prI
```

\newpage

```{r, warning=FALSE, message=FALSE, size="small"}
# Base quality plot
prII <- plotQualityProfile(fnFs[3:4])
prII
```

\newpage

```{r, warning=FALSE, message=FALSE, size="small"}
# Base quality plot
prIII <- plotQualityProfile(fnFs[5:6])
prIII
```

\newpage

Filtering reads (maxEE ≈ 1 error/200 bp sequence)

```{r, warning=FALSE, message=FALSE, size="small"}
# Filtered files are placed in filtered subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
# For single end reads without phix control
names(filtFs) <- sample.names
out <- filterAndTrim(fnFs, filtFs, truncLen=truncation,
                     maxN=0, maxEE=7 , truncQ=2,
                     compress=TRUE, multithread=TRUE,
                     rm.phix=FALSE)
```
